Analysing the Performance of Shallow-Water Models with the Roofline Model

### 2016-01-29 docs/papers/roofline_model/gocean_roofline.tex
roofline: update kernel-vect-perf plot and add discussion

In Figure~\ref{FIG_roofline_haswell} we plot the Roofline model
obtained for the E5-1620 CPU. In order to investigate the poor
performance of the u-Momentum kernel of NEMOLite2D, we also show the
performance of the (simpler) Continuity kernel from that code. As we
would expect, the performance of the u-Momentum kernel remains poor on
this CPU. For the $512^2$ domain the Continuity kernel performs well,
reaching the limit imposed by the bandwidth to main memory. However,
the working-set sizes for this kernel for the $128^2$ and $256^2$
domains are approximately 1~MB and 4.2~MB, respectively and therefore
both of these should fit comfortably within L3 cache. Despite this,
the performance achieved with these domain sizes falls short of the L3
bandwidth limit.  Although this kernel is one of several within the
time-stepping loop of NEMOLite2D, the working-set size of this loop
for the $128^2$ domain is just 2.4~MB which again is well within the
capacity of L3.  . Need to look at performance variation with problem
size and maybe HWPCs at the same time.  What does loop1 of Shallow
achieve on smaller problem sizes?

@@

In order to investigate why the performance of the Continuity kernel
is falling well short of the maximum predicted by the roofline model
we have experimented with the level of SIMD
vectorisation. Figure~\ref{FIG_kernel_perf} shows the performance of
the Continuity and u-Momentum kernels as a function of problem size.

@@

>>>>

In Figure~\ref{FIG_roofline_haswell} we plot the Roofline model
obtained for the E5-1620 CPU. In order to investigate the poor
performance of the u-Momentum kernel of NEMOLite2D, we also show the
performance of the (simpler) Continuity kernel from that code. As we
would expect, the performance of the u-Momentum kernel remains poor on
this CPU. For the $512^2$ domain the Continuity kernel performs well,
reaching the limit imposed by the bandwidth to main memory. However,
the working-set sizes for this kernel for the $128^2$ and $256^2$
domains are approximately 1~MB and 4.2~MB, respectively and therefore
both of these should fit comfortably within L3 cache. Despite this,
the performance achieved with these domain sizes falls short of the L3
bandwidth limit.  Although this kernel is one of several within the
time-stepping loop of NEMOLite2D, the working-set size of this loop
for the $128^2$ domain is just 2.4~MB which again is well within the
capacity of L3.

@@

In order to investigate why the performance of the Continuity kernel
is falling well short of the maximum predicted by the roofline model
we have experimented with the level of SIMD
vectorisation. Figure~\ref{FIG_kernel_perf} shows the performance of
the Continuity and u-Momentum kernels when compiled without
vectorisation (``-no-vec''), with SSE4 (Streaming SIMD Extensions)
vectorisation (``-axSSE4.2'') and with AVX (``-xHost''). SSE supports
vector lengths of 128 bits (two double-precision words) and AVX 512
bits (four double-precision words). For the 64, 128 and 256 problem
sizes, changing from no vectorisation to SSE increases the performance
of the Continuity kernel by up to roughly 60\%; some way short of the
ideal speed-up of 100\% that would be obtained with perfect
SSE vectorisation of a compute-bound kernel. Unsurprisingly then, moving
from SSE to AVX vectors does not give any performance improvement and
in fact, is slightly detrimental for most of the problem sizes.

@@

Peak absolute performance is obtained for the 256 problem size with
SSE instructions. The 512 problem size spills from L3 cache and
becomes bound by the bandwidth to main memory: SIMD vectorisation
gives no performance benefit. The fact that best performance is
obtained when the problem size fits within L3 indicates that there are
other overheads that dominate when the problem size is small enough to
fit within L2. This is possibly because the inner, SIMD-vectorised
loop then only has a trip-count of 35. (This is not 32 because of the way
the simulation domain is set-up; it is the {\em internal} region of
the domain that has dimension $32^2$.)

Turning now to the u-Momentum kernel (lower plot in
Figure~\ref{FIG_kernel_perf}), it is striking that the kernel
performance is almost independent of problem size. The reason for this
becomes clear when we edit the kernel to remove the many IF statements
(present because of the application of boundary conditions). The blue
symbols mark the resulting performance which is significantly
increased and again reaches a maximum at the 256 problem size.  It
should be noted that the compiler reported that this loop had been
vectorised, even before the IF statements were removed. Clearly the
vectorisation is more efficient once the code is simplified. However,
even with the up-lift in performance given by removing all the IF
statements, the performance of this kernel is still far short of even
the main-memory bandwidth limit, despite the fact that the $256^2$
problem fits within L3 cache.
